### 2016-01-26 methods.Rmd
ENH:  More work.

* Importance of machine learning in medical image analysis
    * give examples
    * (tools) support vector machines
    * (tools) neural networks
    * (apps)
    * Random forests
        * MS lesion detection (Geremeria 2011)
        * Brain tumor segmentation (BRATS 2012 and BRATS 2013)
        * Tustison 2015 (importance of feature images)
        * Pustina 2016

@@

Machine learning and pattern recognition techniques have proven extremely useful
for various medical image analysis workflows (see, for example, the annual
Workshop on Machine Learning in Medical Imaging held in conjunction with the Medical
Image Computing and Computer-Aided Intervention international meeting [@mlmi2015]).
Popular techniques such as support vector machines and neural networks have been applied
successfully to clinically relevant imaging tasks such as segmentation
(e.g., [@Bauer:2011aa]) and diagnostic prediction (e.g., [@Tong:2014aa;@Liu:2013aa]).

Another popular machine learning technique that has acheieved much success is the
random forest

The random forest machine learning framework provides an excelllent fromework for medical
image segmentation

Supervised methodologies are uniquely characterized, in part, by the feature images that
are used to identify the regions of interest.

Intensity information alone is insufficient for removing false positives.  For example,
as pointed out in [@Neema:2009aa], higher intensities can be found at the periventricular
caps in normal subjects which often confounds automated lesion detection algorithms.

@@

![Representation of Stage 1 feature images for subject 01C1019.  Input for each
subject consists of FLAIR, T1-, and T2-weighted images which are rigidly pre-aligned
[@Avants:2014aa] to the
space of the T1 image.  The three images are then preprocessed (N4 bias correction
[@Tustison:2010ac] and denoising [@Manjon:2010aa]) followed by application of standard ANTs
brain extraction and $n$-tissue segmentation protocols using the MMRR
symmetric template corresponding and priors [@Tustison:2014ab] to the T1 image.  The
feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right which gives a probabilistic classification of
tissue type.
](Figures/featureImages.png)

![Average \texttt{MeanDecreaseAccuracy} plots generated from the creation of all 24 random
forest models for both Stage 1 and Stage 2 during the leave-one-out evaluation.  These plots
are useful in providing a quantitative assessment of the predictive importance of each feature.
The error bars provide the 95^{th} percentile (i.e., $1.96 \times sigma$) and illustrate the
stability of the feature importance across models.](Figures/averageLeaveOneOut.png)

>>>>

* Importance of machine learning in medical image analysis
    x give examples
    x (tools) support vector machines
    x (tools) neural networks
    x (apps)
    x Random forests algorithms
        x MS lesion detection (Geremeria 2011)
        x Brain tumor segmentation (BRATS 2012 and BRATS 2013)
        x Tustison 2015 (importance of feature images)
        x Pustina 2016
    x Random forests overview

@@

### Random forests for machine learning

Machine learning and pattern recognition techniques have seen increased application
for various medical image analysis workflows (see, for example, the annual
Workshop on Machine Learning in Medical Imaging held in conjunction with the Medical
Image Computing and Computer-Aided Intervention international meeting [@mlmi2015]).
Popular techniques such as support vector machines and neural networks have been applied
successfully to clinically relevant imaging tasks such as supervised image segmentation
(e.g., [@Bauer:2011aa]) and diagnostic prediction (e.g., [@Tong:2014aa;@Liu:2013aa]).
Facilitating the current employment of such techniques are the number of available imaging
data sets [@Van-Horn:2014aa] and the public availability of data science packages such as
SciPy [@scipy] and the R project for statistical computing [@R] and their associated add-on
toolkits.

Random forests [@breiman2001] is a popular machine learning technique that has
demonstrated significant utitility for supervised segmentation tasks (e.g.,
normal human brain segmentation) and other computer vision applications
(e.g., human gait detection [@viola2005]).
In the context of neuropathology, random forest-based paradigms have been employed in the
delineation of multiple sclerosis lesions [@geremia2011], stroke lesions [@Pustina:2016aa],
and brain tumors [@geremia2012;@bauer2012;@zikic2012;@Tustison:2014aa].  Of note, these
latter random forest approaches for brain tumor segmentation have performed well in recent
international competitions.  In response to the lack of objective comparisons between
segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in
2012 [@Menze:2015aa] and has continued every year since under the auspices of the International Conference
of Medical Image Computing and Computer Assisted Interventions (MICCAI).

Random forests are conceptually simple [@breiman2001].  They consist of ensembles of decision trees
that are built from training data.  Once constructed, data to be classified is "pushed"
through each decision tree resulting in a single classification "vote" per tree.  These
votes are then be used for regression or classification of the data.  Although decision
trees had been extensively studied previously, the success of employing collections of
such weak learners for boosting machine learning performance
(e.g., AdaBoost [@schapire1990;@freund1997]) influenced the similarly sytled conglomeration
of decision trees into "forests" with randomized node optimization [@ho1995;@amit1997].
Finally, Breiman [@breiman2001] improved accuracy by random sampling of the training
data (i.e., "bagging") resulting in the current random forest framework.

Crucial to these supervised segmentation approaches are the creation and selection of
"features" as input in conjunction with the ground-truth for model construction.  For the
targeted application in this work (i.e., white matter hyperintensities),
regression/classification are performed at the voxelwise level.  In other words, each voxel
within the region of interest is sent through the ensemble of decision trees and receives
a set of classification votes from each tree permitting a regression or classification
solution.  Since this procedure is performed at the voxelwise level, intensity information
alone is insufficient for good segmentation performance since it lacks spatial context.
For example, as pointed out in [@Neema:2009aa], higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated lesion detection
algorithms.  Other potential confounds include MR signal inhomogeneity and noise.  Therefore,
even though machine learning and pattern recognition techniques are extremely powerful and
have significant potential, just as crucial to outcome is the creative construction and
deployment of salient feature images which we detail below.

### Feature images for WMH segmentation

@@

Supervised methodologies are uniquely characterized, in part, by the feature images that
are used to identify the regions of interest.  In Table 1, we provide a list and basic
categorization of the feature images used for the initial (i.e., Stage 1---more on the use
of multiple random forest stages below) segmentation of the white matter hyperintensities.
In addition Figure 1 provides a representation of a set of feature images for a single
subject analyzed in this work.

![Representation of Stage 1 feature images for subject 01C1019.  The
FLAIR, T1-, and T2-weighted images are rigidly pre-aligned
[@Avants:2014aa] to the space of the T1 image.  The three images are then preprocessed
(N4 bias correction [@Tustison:2010ac] and adaptive denoising [@Manjon:2010aa]) followed by
application of standard ANTs brain extraction and $n$-tissue segmentation protocols using
the MMRR symmetric template corresponding and priors [@Tustison:2014ab] to the T1 image.
The feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right which gives a probabilistic classification of
tissue type.  Not shown are the probability and voting images for the brain stem and
cerebellum.](Figures/featureImages.png)

As mentioned previously, input for each subject comprises FLAIR, T1-, and T2-weighted
acquisitions.  The FLAIR and T2 images are rigidly registered to the T1 image using
the open-source Advanced Normalization Tools (ANTs) [@Avants:2014aa].  The aligned images
are then preprocessed using the denoising algorithm of [@Manjon:2010aa] followed by
N4 bias correction [@Tustison:2010ac] which are then normalized to the intensity
range $[0,1]$.  Although we could have used the intensity standardization algorithm of
[@nyul2000], we found that a simple linear rescaling produced better results.

The T1 image is then processed via the ANTs brain
extraction and tissue segmentation protocols described in [@Tustison:2014ab] in order to
produce a mask for the brain parenchyma and provide probabilistic estimates of the
cerebrospinal fluid (csf), gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  Segmentation is performed using the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2014aa] created from the public
MMRR data set  [@landman2011].

To model the intensity information the first set of images simply includes the
preprocessed and normalized intensity FLAIR, T1, and T2 images.  We also calculate a set
of neighborhood statistics (mean, standard deviation, and skewness) feature images using
a radius of one voxel. For each of the normalized
images, we calculate the difference in intensities with the corresponding warped template
component.  Previous success in the international brain tumor segmentation
competition [@Menze:2015aa] was based on an important set of intensity features that were created
from multi-modal templates mentioned previously [@Tustison:2014aa].  We employ the
same strategy here.  For example, the template
difference feature image for the FLAIR image, $S_{FLAIR}$ is calculated as:

$$S_{FLAIR} - T_{FLAIR}\left(\phi_b^{-1}\right)$$

where $\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$ is the transform
which maps from the individual subject space to the
template space and $T_{FLAIR}$ is the FLAIR template component.  Also,
to take advantage of the bilateral symmetry of the normal brain (in terms of both shape
and intensity), and the fact that the presence of WMH violates that assumption,
we use the symmetric templates to compute the contralateral intensity
differences as an additional intensity feature.  For the FLAIR component, this contralateral
difference image is calculated from

$$S_{FLAIR} - S_{FLAIR}\left(\phi_b^{-1}\left(\phi_R\left(\phi_b\right)\right)\right)$$

where $\phi_R$ denotes a horizontal reflection perpendicular to the mid-sagittal plane of
the symmetric template.

The segmentation probability images described above are used as feature images to provide a
spatial context for the random forest model prediction step.  Additional spatial contextual
feature images include the distance maps [@maurer2003] based on the csf, gray matter, and
deep gray matter images.  These latter images are intended to help distinguish white matter
hyperintensities from false positives induced by the partial voluming at the gray/white
matter interface.  A third set of images are based on the voxel location within the
space of the template.  The T1 image of the subject is registered to the T1 template
component using a B-spline variant [@Tustison:2013ac] of the well-known ANTs Symmetric
Normalization (SyN) algorithm [@Avants:2011ab].  Since the inverse transform is also
derived as part of the registration process, we then register the voxel index locations
back to the space of the individual subject.  These index images are then warped back to
the space of the individual subject.  Note that this is similar in motivation to the work
of [@Anbeek:2004aa].  However, this previous work lacks the normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.

### Stacked/cascaded/concatenated random forests for improved segmentation performance

It has been previously observed that

### Code and data availability

### Evaluation protocol

# Results

![Average \texttt{MeanDecreaseAccuracy} plots generated from the creation of all 24 random
forest models for both Stage 1 and Stage 2 during the leave-one-out evaluation.  These plots
are useful in providing a quantitative assessment of the predictive importance of each feature.
The error bars provide the $95^{th}$ percentile (i.e., $1.96 \times \sigma$) and illustrate the
stability of the feature importance across the leave-one-out models.](Figures/averageLeaveOneOut.png)
